---
title: "Allele Frequencies Net Database"
author: "Kamil Slowikowski"
date: "`r format(Sys.Date())`"
output:
  github_document:
    toc: true
    toc_depth: 2
  html_document:
    self_contained: true
---

```{r,include=FALSE}
options(width=100)
library(data.table)
library(dplyr)
library(glue)
library(readr)
library(magrittr)
library(ggplot2)
library(ggstance)
```

## Introduction

Here, we share a script `allelefrequencies.py` to download allele frequencies
for HLA, KIR, MIC, and cytokine genes from [Allele Frequency Net
Database](http://allelefrequencies.net) (AFND).

We also share [`r length(Sys.glob("allelefrequencies.net/*/*.tsv"))` TSV
files](allelefrequencies.net/) ready for analysis.

Each file looks like this:

```{r}
d <- fread("allelefrequencies.net/hla/A.tsv")
head(d)
```

Definitions:

```
alleles_over_2n (Alleles / 2n)
Allele Frequency: total number of copies of the allele in the population sample
in three decimal format.

indivs_over_n (Individuals / n)
Percentage of individuals who have the allele or gene.
```

## Examples

Here are a few examples of how we can use R to analyze these data.

View the largest and smallest populations available in the data:

```{r}
d %>%
  mutate(n = parse_number(n)) %>%
  select(population, n) %>%
  unique() %>%
  arrange(-n)
```

Sum the allele frequencies for each population and see if they add up to 100%:

```{r}
d %>%
  mutate(alleles_over_2n = parse_number(alleles_over_2n)) %>%
  group_by(population) %>%
  summarize(sum = sum(alleles_over_2n)) %>%
  count(sum == 1)
```

```{r, include = FALSE}
theme_set(
  theme_bw(base_size = 14) +
  theme(
    plot.caption.position = "plot"
  )  
)
```

Plot the frequency of a specific allele in populations with more than 1000
sampled individuals:

```{r, dpi = 300, fig.width = 9, fig.height = 7}

d <- fread("allelefrequencies.net/hla/DQB1.tsv")
my_allele <- "DQB1*02:01"
my_d <- d %>% filter(allele == my_allele) %>%
  mutate(
    n = parse_number(n),
    alleles_over_2n = parse_number(alleles_over_2n)
  ) %>%
  filter(n > 1000) %>%
  arrange(-alleles_over_2n)

ggplot(my_d) +
  aes(x = alleles_over_2n, y = reorder(population, alleles_over_2n)) +
  scale_y_discrete(position = "right") +
  geom_colh() +
  labs(
    x = "Allele Frequency (Alleles / 2N)",
    y = NULL,
    title =  glue("Frequency of {my_allele} across populations"),
    caption = "Data from AFND http://allelefrequencies.net"
  )
```

## Citation

If you use this data, please cite the latest manuscript about **Allele Frequency
Net Database**:

- Gonzalez-Galarza FF, McCabe A, Santos EJMD, Jones J, Takeshita L,
  Ortega-Rivera ND, et al. [Allele frequency net database (AFND) 2020 update:
  gold-standard data classification, open access genotype data and new query
  tools.](https://pubmed.ncbi.nlm.nih.gov/31722398) Nucleic Acids Res. 2020;48:
  D783–D788. doi:10.1093/nar/gkz1029

Also, I want to say thanks to David A. Wells for sharing [scrapeAF][1], which
inspired me to work on this project.

[1]: https://github.com/DAWells/scrapeAF
